Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-25 tex/classical.tex
Rewording: sec. 6 (disc. and concl).
Rewording: All sections finished but 5.

Boltzmann Kinetic theory of gases is one of the theoretical basese of our present 
knowledge of out equilibrium systems. This represent a fundamental connection
between the microscopic nature of matter and the phenomenological macroscopic  laws of classical thermodynamics.
The stochasticity introduced in the, in other way deterministic, kinetics of the
particles of the gas, via the molecular chaos hypothesis, allows the demostration 
of the celebreted Boltzmann $H$ theorem. In this article we propose an alternative approach, which is 
developed by means of a variational method, applied to a functional.
Let us start this section by briefly remaind of the important elements of the 
standard derivation and demostration of the Boltzmann transport equation and the $H$ theorem as they are presented in classical textbooks \cite{bib:huang}.

@@

Boltzmann kinetic theory of gases starts by defining the distribution function 
$f(\vec{r},\vec{v},t)$ as the average number of molecules that at time $t$, has the 
values of position and velocity in the element of volumen in the $\mu$ space.
Assuming a deterministic Newtonian description of the movement of the molecules 
and the invariance of the volume measure of the $\mu$ space, one arrives to the Boltzmann rate equation.
%
\begin{equation}
	f(\vec{r}+\vec{v}\delta t, \vec{v}+\vec{F} \delta t, t+\delta t)=f(\vec{r},\vec{v},t)+\left( \frac{\partial f}{\partial t} \right)_{coll} \delta t.
\end{equation}
%
The collision term $\left( \frac{\partial f}{\partial t}\right)$ of this equation 
describes the fuxes in and out from the volume element due to the collisions.
The integro-differential Boltzmann transpor equation can be cast into.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(\tilde f_2\tilde f_1-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($\tilde f_1$ and $\tilde f_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the functional \cite{bib:tolman,bib:huang}
%
\begin{equation}\label{eq:hbfunctional}
   H_{B}=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Notice that $f(\vec{v},t)$ is a spacially homogeneous distribution function.
The functional $H_B$,
originally introduced by Boltzmann in 1872,
describes a diluted gas, at temperature $T$, with a total energy $E$, a total
number of free classical particles $N$, and occupying a volume $V$.
In order to clearly distinguish the Boltzmann functional $H_B$, we denote here
the Maxwell-Boltzmann distribution function as $f_B$.

The physically correct espontaneous time evolution of an out of equilibrium dilute gas 
is corroborated by the $H$ theorem.
This theorem establishes that if the homogeneous function $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and consequently, if the molecular chaos hypothesis is valid, then
the system evolves in such a way that $dH_B/dt\leq0$, and if when $dH_B/dt=0$, 
the system is in the equilibrium state. $H$-theorem is easily proven by 
using the Boltzmann transport equation (\ref{eq:transport}) \cite{bib:huang}. 
$H$-theorem assures that our microscopic approach to describe the spontaneous 
time evolution of the system will be consistent with the phenomenological observations
established in the 2nd-law of Classical; in fact,
$H_B$ can be associated with an entropy density.

On the other hand, considering a diluted gas in equilibrium, \textit{i.e.}
$(\partial f/\partial t)=0$,
with no applied external forces $f$ is independent of $\vec r$, 
it can be directly proven that the equilibrium distribution function obtained from
Eq.~(\ref{eq:transport}) is precisely the Maxwell-Boltzmann
distribution function. The proof of the above commences by finding the sufficient
condition for $f$ to render a null r.h.s. of Eq. (\ref{eq:transport}). Such an
$f$, which we denote here as $f_0$, must satisfy:
%
\begin{equation}\label{eq:neccondf0}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%
Subsequently, the Maxwell-Boltzmann distrubution function can be obtained by
taking the logarithm to Eq.~(\ref{eq:neccondf0}), and exploiting mechanical
conserved quantities (see \cite[ch. 4.2]{bib:huang}).

Before introducing our proposal, we call the reader's attention to the following. In defining
$H_B$ Eq.~(\ref{eq:hbfunctional}), it is assumed that the distribution function 
$f_B$ is spattially homogeneous. Clearly this assumption simplifies the demostration 
of the $H$ theorem, nevertheless, it introduces as well a limitate-conception
of an out of equilibrium condition in the gas. Given the relatively simple nature of a dilute 
gas, one of the main characteristics of an out of equilibrium condition would be the existence of
inhomogeneities in the system.

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is fullfiled. In order to extend the procedure for systems whose
distribution functions are not necessarily homogeneous, would possibilities allows 
for studying systems been in a more general out of equilibrium condition, we introduce a modified $H$-functional 
as follows. For the sake of clearity and simplicity we will use primed functions
and quantities to denotate the classical case and differentiate it from the quantum
case.

Our proposed functional, denoted as $\mathcal{H}'$, describes a dilute classical
gas occupying a volume $V$. For our theoretical treatment we will consider this volume
divided into $K$ cells, which, without loss of generality,
have identical volumes $\delta V_M = V/K,\ M=1,\,\dots\,,K$.
Locally the $\mathcal{H}'$-functional is denoted by $\mathcal{H}'_M$, and 
the local homogeneous distribution function, $f'_{M}(\vec{v},t)$; local 
number of particles, $\mathcal{N}'_M$; local temperature, $T'_M$; and a local 
energy, $\mathcal{E}'_M$. As a whole, the system has a
energy $E$, and a global number of free classical particles $N$. Let us assume that the system is 
to be isolated. We also assume that the number of particles
in each cell is sufficiently large, so as to obtain accurate averages.
We start our analysis by proposing the following inhomogeneous functional
%
\begin{equation}\label{eq:cHdef}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int_{\delta V_M} f'_M(\vec{v},t) \ln f'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

The distribution functions, $\{f'_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Notice that each $f'_M$ can be formally extended to the complete coordinate space
by defining each $f'_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f'_M(\vec v,t)\}$:
%
\begin{equation}
   f'(\vec r,\vec v,t)=\sum_{M=1}^Kf'(\vec r_M,\vec v,t)=\sum_{M=1}^Kf'_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f'_M(\vec v,t)\neq f'_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f'_M(\vec{v},t)$ and a local variable of energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    \mathcal{H}_M' & = &  \int f'_M(\vec{v},t) \ln f'_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f'_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f'_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming homogeneous distribution functions per cell means that we are accepting valid the 
local equilibrium hypothesis. In addition, the set $\{f'_{M}(\vec{v},t)\}$ must satisfy the following restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

We now use the variational method in order to find the extremal of $\mathcal{H}'$, consistent with the restrictions
(\ref{eq:cellrestrictions}) and (\ref{eq:micro}) by using the corresponding Lagrange
multipliers $\{\alpha_M\}$ and $\{\beta_M\}$. This yields
%
\begin{eqnarray}\label{eq:deltaHpdeltafpj}
    \frac{\delta \mathcal{H}'}{\delta f'_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f'_J(\vec{v}')}\left[
        f'_M(\vec{v})\ln f'_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f'_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f'_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f'_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f'_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that the use of the variational procedure on
$\mathcal{H}'$ predicts that when the equilibrium is reached, the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, on a statistical sense, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f'_M$
does not depend on the cell number $M$ (\textit{i.e.} $f'_M(\vec v,t)=f'(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$. Here we have used Eqs. (\ref{eq:Ecell}) and (\ref{eq:Ncell}).
Eq.~(\ref{eq:cHdef}), in terms of Eq.~(\ref{eq:hbfunctional}), can be cast into
%
\begin{equation}
    \mathcal{H}'(t)=\int \sum_{M=1}^{K} [f'(\vec{v},t)\ln f'(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $\mathcal{H}'$ with the entropy, still we need to show that $\mathcal{H}'(t)$ is extensive,
with respect to $Kf'_(\vec v,t)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f'(\vec{v},t) + Kf'(\vec{v},t) \ln f'(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f'(\vec{v},t)\big[\ln K + \ln f'(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f'(\vec{v},t)$, is much large than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v \simeq
    K\int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $\mathcal{H}'_M$ is extensive, and the sum $\sum_M\mathcal{H}'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional.
\commb{Hence, $\mathcal{H}'$ can also be identified with a density of entropy for the system.}

Furthermore, in equilibrium the Lagrange multipliers are position- and time-independent,
thus $f'_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f'_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}'(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $\mathcal{H}''$
functional:
%
\begin{equation}
   \mathcal{H}''(t)=\sum_{M=1}^{K}\int \left[f'_M(\vec{v},t)
    \ln f'_M(\vec{v},t)-f'_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
since $\sum_{M=1}^{K} \int f'_M(\vec{v},t)\mathrm{d}^3v =N$.
$N$ is a constant and we are mainly intersted in the time-derivative of $\mathcal{H}''$, thus
$C$ can be conveniently omitted. In other words, $\mathcal{H}'$ leads the Maxwell-Boltzmann
distribution function for systems in equilibrium.

@@

We can observe that $\bar{f}(\vec{v})$ in (\ref{eq:classicalH4}) is always
positive. Also the deviation and its derivative are positive, then all the
expression is positive. However, the global sign makes the derivative of the
$\mathcal{H}'$ with respect to time is always less to zero. With this we proved that
$\frac{d\mathcal{H}'}{dt}<0$. If the system is in equilibrium,
$g_{J}(\vec{v},t)=\dot g_J(\vec{v},t)=0$, therefore $\frac{d\mathcal{H}'}{dt}=0$.
\end{comment}
%-----------------------------
%
If the system's initial state is sufficiently close to equilibrium, it is
physically suitable
to assume that $\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$ in a monotonous manner,
thus ${{\dot{g}}'}_I\phantom{.}\!^{+}(\vec{v},t)\leq0$ and
${{\dot{g}}'}_J\phantom{.}\!^{-}(\vec{v},t)>0$, for $t\geq t_0$. Consequently,
Eq.~(\ref{eq:classicalH3}) can be re-written as
%
\begin{equation}\label{eq:classicalH4}
    \frac{d\mathcal{H}'}{dt}=-\left[
      \sum_J^{L}\int\bar f(\vec{v})|g_J^{+}(\vec{v},t)|
        |\dot g_J^{+}(\vec{v},t)|\mathrm{d}^3v
      +\sum_J^{P}\int\bar f(\vec{v})|g_J^{-}(\vec{v},t)|
      |\dot g_J^{-}(\vec{v},t)|\mathrm{d}^3v 
    \right].
\end{equation}
%
Since every integrand in Eq.~(\ref{eq:classicalH4}) is positive,
for all $t$ and $\vec v$, and
$\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$, it follows that
%
\begin{equation}\label{eq:dHpdtleq0}
 \frac{d\mathcal{H}'}{dt}\leq0.
\end{equation}
%

>>>>

Boltzmann Kinetic theory of gases represents a fundamental connection
between the microscopic nature of matter and the phenomenological macroscopic  laws of classical thermodynamics.
The stochasticity introduced by the molecular chaos hypothesis in the, otherwise deterministic, kinetics of the
particles allows the demostration 
of the celebrated Boltzmann $H$-theorem. In contrast, in this article we propose an alternative approach, which is 
developed by means of a variational procedure applied to an $H$-functional.
Let us start this section by briefly accounting for the important elements of the 
standard derivation of the Boltzmann transport equation and demostration the $H$-theorem,
such as they are presented in classical textbooks \cite{bib:huang}.

@@

In Boltzmann kinetic theory of gases, one starts by defining the distribution function, 
$f(\vec{r},\vec{v},t)$, as the average number of molecules that, at time $t$, have 
position $\vec r$ and velocity $\vec v$, and are contained in a $\mu$-space volume element $d^3rd^3v$.
Assuming a deterministic Newtonian description of the molecules motion, 
as well as the invariance of the $\mu$-space volume measure, one arrives to the Boltzmann rate equation:
%
\begin{equation}
	f(\vec{r}+\vec{v}\delta t, \vec{v}+\vec{F} \delta t, t+\delta t)=f(\vec{r},\vec{v},t)+\left( \frac{\partial f}{\partial t} \right)_{\textrm{coll}} \delta t.
\end{equation}
%
Here, the term $\left(\partial f/\partial t\right)_{\textrm{coll}}$
describes the in and out fluxes from and towards the volume element, due to the collisions.
Subsequently, from the previous equation, the integro-differential Boltzmann transport equation is obtained:
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(\tilde f_2\tilde f_1-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($\tilde f_1$ and $\tilde f_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the $H$-functional \cite{bib:tolman,bib:huang}:
%
\begin{equation}\label{eq:hbfunctional}
   H_{B}=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Notice that $f_B(\vec{v},t)$ is a spatially homogeneous distribution function.
The functional $H_B$,
originally introduced by Boltzmann in 1872,
describes a diluted gas occupying a volume $V$, at temperature $T$, with total energy $E$, and total
number of free classical particles $N$.
In order to clearly distinguish the Boltzmann functional $H_B$, we denote hereafter
the Maxwell-Boltzmann distribution function as $f_B$.

The physically correct espontaneous time-evolution of an out of equilibrium diluted gas 
is corroborated by the $H$-theorem.
This theorem establishes that if the two conditions, a) the homogeneous
function $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and b) the molecular chaos hypothesis is valid, then
the system evolves in such a manner that $dH_B/dt\leq0$, and if $dH_B/dt=0$, 
then the system is in the equilibrium state. 
The $H$-theorem is straightforward to prove through Eq.~(\ref{eq:transport}) \cite{bib:huang}, 
and it assures the consistency between our microscopic approach to describe the system's spontaneous 
time-evolution and the phenomenological observations
established by the $2^{\textrm{nd}}$ law of classical thermodynamics; in fact,
$H_B$ can be associated with an entropy density.

On the other hand, considering a diluted gas in equilibrium
with no applied external forces, \textit{i.e.}
$(\partial f/\partial t)=0$ and $f$ is independent of $\vec r$), 
it can be directly proven that the equilibrium distribution function obtained from
Eq.~(\ref{eq:transport}) is precisely the Maxwell-Boltzmann
distribution function. The proof of the above commences by finding the sufficient
condition for $f$ to render a null r.h.s. of Eq. (\ref{eq:transport}). Such an
$f$, which we denote here as $f_0$, must satisfy:
%
\begin{equation}\label{eq:neccondf0}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%
Subsequently, the Maxwell-Boltzmann distrubution function can be obtained by
taking the logarithm of Eq.~(\ref{eq:neccondf0}), and exploiting mechanical
conserved quantities (see \cite[ch. 4.2]{bib:huang}).

Before introducing our proposed $H$-functional, we remark the following. In defining
$H_B$, Eq.~(\ref{eq:hbfunctional}), it is assumed that the distribution function 
$f$ is spatially homogeneous. This assumption simplifies the demostration 
of the $H$ theorem, however, it introduces as well a limited conception
of the out-of-equilibrium condition of the gas. Given the relatively simple nature of a dilute 
gas, one of the salient features of an out-of-equilibrium condition is the existence of
inhomogeneities in the system.

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is fullfiled. In order to extend the previous procedure to systems whose
distribution functions are non-homogeneous, which might allow for studying systems
being in a more general out-of-equilibrium condition, we introduce a modified $H$-functional 
in the following. For the sake of clearity and simplicity, we will use primed functions
and quantities to denote the classical case, so as to differentiate them from the quantum
analogues.

Our proposed classical $H$-functional, denoted as $\mathcal{H}'$, describes a dilute classical
gas occupying a volume $V$. In our theoretical treatment, we divide this volume
into $K$ cells, which, without loss of generality,
have identical volumes, $\delta V_M = V/K,\ M=1,\,\dots\,,K$.
Each cell of index $M$ has the following local functions, properties, and variables:
an $H$-functional, $\mathcal{H}'_M$, a homogeneous distribution function, $f'_{M}(\vec{v},t)$,
a number of particles, $\mathcal{N}'_M$, a temperature, $T'_M$, and an
energy, $\mathcal{E}'_M$. Taken as a whole, the system has a
energy $E$, and a global number of free classical particles $N$. We also assume that the system is 
perfectly isolated, and that the number of particles
in each cell is sufficiently large, so as to obtain accurate averages.
We start our analysis by proposing the following inhomogeneous $H$-functional:
%
\begin{equation}\label{eq:cHdef}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int_{\delta V_M} f'_M(\vec{v},t) \ln f'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

The distribution functions, $\{f'_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Notice that each $f'_M$ can be formally extended to the complete coordinate space
by defining each $f'_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f'_M(\vec v,t)\}$:
%
\begin{equation}
   f'(\vec r,\vec v,t)=\sum_{M=1}^Kf'(\vec r_M,\vec v,t)=\sum_{M=1}^Kf'_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f'_M(\vec v,t)\neq f'_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f'_M(\vec{v},t)$ and a local variable of energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    \mathcal{H}_M' & = &  \int f'_M(\vec{v},t) \ln f'_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f'_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f'_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming that every $f'_M$ is homogeneous implies that we are accepting the validity of the
local equilibrium hypothesis. In addition, the set $\{f'_{M}(\vec{v},t)\}$ must satisfy the following restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

We now use the variational method to find the extremal of $\mathcal{H}'$,
consistent with the restrictions
(\ref{eq:cellrestrictions}) and (\ref{eq:micro}), together with the corresponding Lagrange
multipliers $\{\alpha_M\}$ and $\{\beta_M\}$. This yields:
%
\begin{eqnarray}\label{eq:deltaHpdeltafpj}
    \frac{\delta \mathcal{H}'}{\delta f'_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f'_J(\vec{v}')}\left[
        f'_M(\vec{v})\ln f'_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f'_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f'_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f'_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f'_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that the use of the variational procedure on
$\mathcal{H}'$ predicts that when the equilibrium is reached, the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, in a statistical sense, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the global distribution function is homogeneous, hence $f'_M$
does not depend on the cell number $M$ (\textit{i.e.} $f'_M(\vec v,t)=f'(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$. Here we have used Eqs. (\ref{eq:Ecell}) and (\ref{eq:Ncell}).
Eq.~(\ref{eq:cHdef}), in terms of Eq.~(\ref{eq:hbfunctional}), can be rewritten as:
%
\begin{equation}
    \mathcal{H}'(t)=\int \sum_{M=1}^{K} [f'(\vec{v},t)\ln f'(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%

In order to identify $\mathcal{H}'$ with the entropy, we need to show that $\mathcal{H}'(t)$ is extensive,
with respect to $Kf'(\vec v,t)$.
This is shown by analyzing the following expression:
%
\begin{eqnarray}\label{eq:sum-h}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f'(\vec{v},t) + Kf'(\vec{v},t) \ln f'(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f'(\vec{v},t)\big[\ln K + \ln f'(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f'(\vec{v},t)$, is much larger than
the number of cells, $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v \approx
    K\int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $\mathcal{H}'_M$ is extensive, and the sum $\sum_M\mathcal{H}'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional.
Therefore, $\mathcal{H}'$ can be identified with the entropy density of the system.

Furthermore, in equilibrium, the Lagrange multipliers are position- and time-independent,
thus $f'_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f'_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}'(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $\mathcal{H}''$
functional:
%
\begin{equation}
   \mathcal{H}''(t)=\sum_{M=1}^{K}\int \left[f'_M(\vec{v},t)
    \ln f'_M(\vec{v},t)-f'_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
Since $\sum_{M=1}^{K} \int f'_M(\vec{v},t)\mathrm{d}^3v =N$ (a constant)
and because we are mainly interested in the time-derivative of $\mathcal{H}''$,
$C$ can be conveniently omitted. In other words, $\mathcal{H}'$ leads to the Maxwell-Boltzmann
distribution function of systems in equilibrium.

@@

We can observe that $\bar{f}(\vec{v})$ in (\ref{eq:classicalH4}) is always
positive. Also the deviation and its derivative are positive, then all the
expression is positive. However, the global sign makes the derivative of the
$\mathcal{H}'$ with respect to time is always less to zero. With this we proved that
$\frac{d\mathcal{H}'}{dt}<0$. If the system is in equilibrium,
$g_{J}(\vec{v},t)=\dot g_J(\vec{v},t)=0$, therefore $\frac{d\mathcal{H}'}{dt}=0$.
\end{comment}
%-----------------------------
%
If the system's initial state is sufficiently close to equilibrium, it is
physically suitable
to assume that $\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$ in a monotonous manner,
thus ${{\dot{g}}'}_I\phantom{.}\!^{+}(\vec{v},t)\leq0$ and
${{\dot{g}}'}_J\phantom{.}\!^{-}(\vec{v},t)>0$, for $t\geq t_0$. Consequently,
Eq.~(\ref{eq:classicalH3}) can be re-written as
%
\begin{equation}\label{eq:classicalH4}
    \frac{d\mathcal{H}'}{dt}=-\left[
      \sum_J^{L}\int\bar f(\vec{v})|g_J^{+}(\vec{v},t)|
        |\dot g_J^{+}(\vec{v},t)|\mathrm{d}^3v
      +\sum_J^{P}\int\bar f(\vec{v})|g_J^{-}(\vec{v},t)|
      |\dot g_J^{-}(\vec{v},t)|\mathrm{d}^3v 
    \right].
\end{equation}
%
Since every integrand in Eq.~(\ref{eq:classicalH4}) is positive,
for all $t$ and $\vec v$, and
$\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$, it follows that
%
\begin{equation}\label{eq:dHpdtleq0}
 \frac{d\mathcal{H}'}{dt}\leq0.\qquad\qquad\textrm{QED.}
\end{equation}
%
